For Finitary Induction-Induction, \\ Induction is Enough

### 2020-05-03 paper/paper.tex
make uppercase greek letters italic

Inductive-inductive types (IITs \cite{forsberg-phd}) allow the mutual
definition of a type and a family of types over the first one. IITs
were originally introduced to represent the well-typed syntax of type
theory itself, and the main example is still the Chapman-style
\cite{chapman09eatitself} syntax for a type theory. A minimised version
is the IIT of contexts and types given by the following constructors.
\begin{alignat*}{5}
  & \Con && : \Set \\
  & \Ty && : \Con\ra\Set \\
  & \mathsf{empty} && : \Con \\
  & \mathsf{ext} && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  & \mathsf{U} && : (\Gamma:\Con)\ra\Ty\,\Gamma \\
  & \mathsf{El} && : (\Gamma:\Con)\ra\Ty\,(\mathsf{ext}\,\Gamma\,(\mathsf{U}\,\Gamma))
%  & \mathsf{Pi} && : (\Gamma:\Con)(A:\Ty\,\Gamma)\ra\Ty\,(\mathsf{ext}\,\Gamma\,A)\ra\,\Ty\,\Gamma
\end{alignat*}
This type has two sorts, $\Con$ and $\Ty$. The $\mathsf{ext}$ constructor of
$\Con$ refers to $\Ty$ and the $\Ty$-constructor $\mathsf{U}$ refers to $\Con$,
hence the two sorts have to be defined simultaneously. Moreover, $\Ty$ is
indexed over $\Con$. This precludes a reduction analogous to the reduction of
$\mathsf{isEven}$--$\mathsf{isOdd}$, as we would get a type indexed over
itself. Another unique feature of IITs (which also holds for higher inductive
types \cite{HoTTbook}) is that later constructors can refer to previous
constructors: in our case, $\mathsf{El}$ mentions
$\mathsf{ext}$.
% \[
%   \mathsf{Con?} : (b:\mathsf{Bool})\ra\mathsf{if}\,b\,\mathsf{then}\,\Set\,\mathsf{else}\,(\mathsf{Con?}\,\mathsf{false}\ra\Set),
% \]

The elimination principle for the above IIT has the following two
motives (one for each sort) and four methods (one for each
constructor).
\begin{alignat*}{5}
  & Con^D && : \Con\ra\Set \\
  & Ty^D && : Con^D\,\Gamma\ra\Ty\,\Gamma\ra\Set \\
  & {empty}^D && : Con^D\,\mathsf{empty} \\
  & {ext}^D && : (\Gamma^D:Con^D\,\Gamma)\ra Ty^D\,\Gamma^D\,A\ra Con^D\,(\mathsf{ext}\,\Gamma\,A) \\
  & U^D && : (\Gamma^D:Con^D\,\Gamma)\ra Ty^D\,\Gamma^D\,(\mathsf{U}\,\Gamma) \\
  & El^D && : (\Gamma^D:Con^D\,\Gamma)\ra\,Ty^D\,({ext}^D\,\Gamma^D\,(U^D\,\Gamma^D))\,(\mathsf{El}\,\Gamma)
%  & Pi^D && : (\Gamma^D:Con^D\,\Gamma)(A^D:\Ty^D\,\Gamma^D\,A)\ra \Ty^D\,(\mathsf{ext}^D\,\Gamma^D\,A^D)\ra\,\Ty^D\,\Gamma^D\,(\mathsf{Pi}\,\Gamma\,A\,B)
\end{alignat*} xxx
Above we used implicit quantifications for $\Gamma:\Con$ and
$A:\Ty\,\Gamma$ to ease readability, e.g.\ $Ty^D$ has an implicit
parameter $\Gamma$ before its explicit parameter of type
$Con^D\,\Gamma$.

Given the above motives and methods the elimination principle provides
two functions
\begin{alignat*}{5}
  & \mathsf{elimCon} && : (\Gamma:\Con)\ra Con^D\,\Gamma \\
  & \mathsf{elimTy} && : (A:\Ty\,\Gamma)\ra Ty^D\,(\mathsf{elimCon}\,\Gamma)\,A
\end{alignat*}
with the following computation rules.
\begin{alignat*}{5}
  & \mathsf{elimCon}\,\mathsf{empty} && = {empty}^D \\
  & \mathsf{elimCon}\,(\mathsf{ext}\,\Gamma\,A) && = {ext}^D\,(\mathsf{elimCon}\,\Gamma)\,(\mathsf{elimTy}\,A) \\
  & \mathsf{elimTy}\,(\mathsf{U}\,\Gamma) && = U^D\,(\mathsf{elimCon}\,\Gamma) \\
  & \mathsf{elimTy}\,(\mathsf{El}\,\Gamma) && = El^D\,(\mathsf{elimCon}\,\Gamma)
\end{alignat*}
The functions $\mathsf{elimCon}$ and $\mathsf{elimTy}$ are an example
of a \emph{recursive-recursive} definition (using nomenclature from
\cite{forsberg-phd}). This means two mutually defined functions where
the type of the second function depends on the first function. The
proof assistant Agda \cite{norell07thesis} allows defining such
functions (even from non-IITs) and it is currently the only proof
assistant supporting IITs\footnote{An experimental version of Coq with
  IITs is also available on GitHub.}.

Reducing IITs to inductive types (more precisely, to indexed W-types)
is an open problem. Forsberg \cite{forsberg-phd} showed a reduction in
extensional type theory however this only provides a simpler,
non-recursive-recursive elimination principle. Hugunin \cite{jasper}
reduced several IITs to inductive types, working inside a cubical type
theory, but he also only constructed the simple eliminator. To
illustrate the difference, we list the motives and methods and the
simple elimination principle for the $\Con$--$\Ty$ example. Again, we
use implicit quantifications.
\begin{alignat*}{5}
  & Con^S && : \Con\ra\Set \\
  & Ty^S && : \Ty\,\Gamma\ra\Set \\
  & {empty}^S && : Con^S\,\mathsf{empty} \\
  & {ext}^S && : Con^S\,\Gamma\ra Ty^S\,\,A\ra Con^S\,(\mathsf{ext}\,\Gamma\,A) \\
  & U^S && : Con^S\,\Gamma\ra Ty^S\,(\mathsf{U}\,\Gamma) \\
  & El^S && : Con^S\,\Gamma\ra\,Ty^S\,(\mathsf{El}\,\Gamma) \\
  & \mathsf{selimCon} && : (\Gamma:\Con)\ra Con^S\,\Gamma \\
  & \mathsf{selimTy} && : (A:\Ty\,\Gamma)\ra Ty^S\,A
\end{alignat*}
This simple elimination principle is too weak e.g.\ to define the
standard (metacircular) interpretation of our small syntax
\cite{ttintt}. Using pattern matching notation, this interpretation is
the following:
\begin{alignat*}{5}
  & \ll\blank\rr && : \Con\ra\Set_1 \\
  & \ll\blank\rr && : \ll\Gamma\rr\ra\Set_1 \\
  & \ll\mathsf{empty}\rr && := \top \\
  & \ll\mathsf{ext}\,\Gamma\,A\rr && := (\gamma:\ll\Gamma\rr)\times\ll A\rr\,\gamma \\
  & \ll\mathsf{U}\,\Gamma\rr\,\gamma && := \Set \\
  & \ll\mathsf{El}\,\Gamma\rr\,(\gamma,X) && := X
\end{alignat*}
The reason that we need the general elimination principle is that
$\ll\blank\rr$ for types refers to $\ll\blank\rr$ for contexts, hence
this function is recursive-recursive.

@@

Just as ibid., we only consider finitary IITs, that is, constructors
can only have a finite number of recursive arguments. An example
constructor for $\Con$--$\Ty$ which is not allowed is the following:
\[
  \Pi_\infty : (\Gamma : \Con) \rightarrow (\mathbb{N} \rightarrow \Ty\,\Gamma) \rightarrow \Ty\,\Gamma
\]

@@

\begin{definition}[Signature algebra, $\SignAlg$]\label{def:algebra}
  In a model of ETT, a signature algebra is an iterated $\Sigma$ type
  consisting of the following four (families of) sets, 17 operations
  and 18 equalities.
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\
  & \Con && : \Set \\
  & \Ty  && : \Con\ra\Set \\
  & \Sub  && : \Con\ra\Con\ra\Set \\
  & \Tm  && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Set \\
  & \id && : \Sub\,\Gamma\,\Gamma \\
  & \blank\circ\blank && : \Sub\,\Theta\,\Delta\ra\Sub\,\Gamma\,\Theta\ra\Sub\,\Gamma\,\Delta \\
  & \ass && : (\sigma \circ \delta) \circ \nu = \sigma \circ (\delta \circ \nu) \\
  & \idl && : \id\circ\sigma = \sigma \\
  & \idr && : \sigma\circ\id = \sigma \\
  & \blank[\blank] && : \Ty\,\Delta\ra\Sub\,\Gamma\,\Delta\ra\Ty\,\Gamma \\
  & \blank[\blank] && : \Tm\,\Delta\,A\ra(\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma]) \\
  & [\id] && : A[\id] = A \\
  & [\circ] && : A[\sigma \circ \delta] = A[\sigma][\delta] \\
  & [\id] && : t[\id] = t \\
  & [\circ] && : t[\sigma \circ \delta] = t[\sigma][\delta] \\
  & \boldsymbol{\cdot} && : \Con \\
  & \epsilon && : \Sub\,\Gamma\,\boldsymbol{\cdot} \\
  & {\boldsymbol{\cdot}\eta} && : (\sigma : \Sub\,\Gamma\,\boldsymbol{\cdot}) \ra \sigma = \epsilon \\
  & \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  & \blank,\blank && : (\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma])\ra\Sub\,\Gamma\,(\Delta\ext A) \\
  & \pi_1 && : \Sub\,\Gamma\,(\Delta\ext A)\ra \Sub\,\Gamma\,\Delta \\
  & \pi_2 && : (\sigma : \Sub\,\Gamma\,(\Delta\ext A))\ra \Tm\,\Gamma\,(A[\pi_1 \sigma]) \\
  & {\pi_1\beta} && : \pi_1(\sigma, t) = \sigma \\
  & {\pi_2\beta} && : \pi_2(\sigma, t) = t \\
  & {\pi\eta} && : (\pi_1\,\sigma, \pi_2\,\sigma) = \sigma \\
  & {,\circ} && : (\sigma, t)\circ\delta = (\sigma\circ\delta, t[\delta]) \\
  & \rlap{(2) Universe} \\
  & \U && : \Ty\,\Gamma \\
  & \El && : \Tm\,\Gamma\,\U\ra\Ty\,\Gamma \\
  & {\U[]} && : \U[\sigma] = \U \\
  & {\El[]} && : (\El\,a)[\sigma] = \El\,(a[\sigma]) \\
  & \rlap{(3) Inductive parameters} \\
  & \Pi && : (a:\Tm\,\Gamma\,\U)\ra\Ty\,(\Gamma\ext\El\,a)\ra\Ty\,\Gamma \\
  & \blank\oldapp\blank && : \Tm\,\Gamma\,(\Pi\,a\,B)\ra (u : \Tm\,\Gamma\,(\El\,a))
  \ra \Tm\,\Gamma\,(\El\,(B[\id,\,u])) \\
  & {\Pi[]} && : (\Pi\,a\,B)[\sigma] = \Pi\,(a[\sigma])\,(B[\sigma\circ\p,\q]) \\
  & {\oldapp[]} && : (t\oldapp\alpha)[\sigma] = (t[\sigma])\mathop{\oldapp}(\alpha[\sigma]) \\
  & \rlap{(4) External parameters} \\
  & \Pim && : (T:\Set)\ra(T\ra\Ty\,\Gamma)\ra\Ty\,\Gamma \\
  & \blank\appm \blank && : \Tm\,\Gamma\,(\Pim\,T\,B)\ra(\alpha:T) \ra\Tm\,\Gamma\,(B\,\alpha) \\
  & {\Pim[]} && : (\Pim\,T\,B)[\sigma] = \Pim\,T\,(\lambda \alpha.(B\,\alpha)[\sigma]) \\
  & {\appm[]} && : (t\appm\alpha)[\sigma] = (t[\sigma])\mathop{\appm}\alpha
\end{alignat*}
Given an $M:\SignAlg$, we denote its components by $\Con^M$, $\Ty^M$,
$\Sub^M$, $\Tm^M$, $\id^M$, and so on. We omit the indices if there is
only one signature algebra in scope (e.g.\ in
\Cref{def:abbrevs,ex:contexts}).
\end{definition}

\begin{definition}[Abbreviations]\label{def:abbrevs}
  For a signature algebra, we use $\wk : \Sub\,(\Gamma\ext A)\,\Gamma$
  to mean $\pi_1\,\id$. We recover de Bruijn indices by setting
  $0:= \pi_2 \,\id$ and $1+n := n[\wk]$. $\Pi\,a\,(B[\wk])$ is
  abbreviated by $a\Ra B$, $\Pim\,T\,(\lambda\_.B)$ by $T\Ram B$.
\end{definition}

@@

  The $\Con$--$\Ty$ example from \Cref{sec:intro} is given by the
  following context.
  \begin{alignat*}{5}
    & \boldsymbol{\cdot} \,\ext                                                 && \boldsymbol{\cdot} \,\ext                                                                \\
    & \U\,\ext                                                                  && Con && : \U\,\ext                                                                        \\
    & 0\Ra\U\,\ext                                                              && Ty && : Con\Ra\U\,\ext                                                                   \\
    & \El\,1\,\ext                                                              && empty && : \El\,Con\,\ext                                                                \\
    & \Pi\,2\,(2\oldapp 0 \Ra \El\,3)\ext                                       && ext && : \Pi\,(\Gamma : Con)\,(Ty\oldapp\Gamma \Ra \El\,Con)\ext                         \\
    & \Pi\,3\,(\El\,(3\oldapp 0))\,\ext                                         && U && : \Pi\,(\Gamma:Con)\,(\El\,(Ty\oldapp\Gamma))\,\ext                                 \\
    & \Pi\,4\,(\El\,(4\oldapp(2\oldapp 0\oldapp (1\oldapp 0)))) \hspace{4.79em} && El && : \Pi\,(\Gamma : Con)\,(\El\,(Ty\oldapp(ext\oldapp\Gamma\oldapp (U\oldapp\Gamma))))
  \end{alignat*}
\end{example}

@@

The notion of morphism is determined by the notion of algebra
\cite{large_inf_qiit}, but we include it here for completeness.
\begin{definition}[Signature morphism, $\SignMor$]\label{def:morphism}
  A morphism from signature algebras $M$ to $N$ denoted
  $\SignMor\,M\,N$ consists of four functions and 17 equalities
  expressing that the functions preserve the operations of the two
  algebras. We use the same naming as in \Cref{def:algebra} and use
  superscripts to denote which algebra is meant.
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\
  & \Con && : \Con^M && \ra\Con^N \\
  & \Ty  && : \Ty^M\,\Gamma && \ra\Ty^N\,(\Con\,\Gamma) \\
  & \Sub  && : \Sub^M\,\Gamma\,\Delta && \ra\Sub^N(\Con\,\Gamma)\,(\Con\,\Delta) \\
  & \Tm  && : \Tm^M\,\Gamma\,A && \ra\Tm^N\,(\Con\,\Gamma)\,(\Ty\,A) \\
  & \id && : \Sub\,\id^M && = \id^N \\
  & {\circ} && : \sigma\circ^M\delta && = \Sub\,\sigma\circ^N\Sub\,\delta \\
  & [] && : A[\sigma]^M && = \Ty\,A[\Sub\,\sigma]^N \\
  & [] && : t[\sigma]^M && = \Tm\,t[\Sub\,\sigma]^N \\
  & \boldsymbol{\cdot} && : \Con\,\boldsymbol{\cdot}^M && = \boldsymbol{\cdot}^N \\
  & \epsilon && : \Sub\,\epsilon^M && = \epsilon^N \\
  & \ext && : \Con\,(\Gamma\ext^M A) && = \Con\,\Gamma\ext^N\Ty\,A \\
  & , && : \Sub\,(\sigma,^M t) && = \Sub\,\sigma ,^N\Tm\,t \\
  & \pi_1 && : \Sub\,(\pi_1^M\,\sigma) && = \pi_1^N\,(\Sub\,\sigma) \\
  & \pi_2 && : \Tm\,(\pi_2^M\,\sigma) && = \pi_2^N\,(\Sub\,\sigma) \\
  & \rlap{(2) Universe} \\
  & \U && : \Ty\,\U^M && = \U^N \\
  & \El && : \Ty\,(\El^M\,a) && = \El^N\,(\Tm\,a) \\
  & \rlap{(3) Inductive parameters} \\
  & \Pi && : \Ty\,(\Pi^M\,a\,B) && = \Pi^N\,(\Tm\,a)\,(\Ty\,B) \\
  & \oldapp && : \Tm\,(t\mathbin{{\oldapp}^M} u) && = \Tm\,t\mathbin{{\oldapp}^N}\,\Tm\,u \\
  & \rlap{(4) External parameters} \\
  & \Pim && : \Ty\,(\Pim^M\,T\,B) && = \Pim^N\,T\,(\lambda\alpha.\Ty\,(B\,\alpha)) \\
  & \appm && : \Tm\,(t\mathbin{{\appm}^M}\alpha) && = \Tm\,t\mathbin{{\appm}^N}\alpha
\end{alignat*}
Given an $f:\SignMor\,M\,N$, we denote its first four components just
by $f_\Con$, $f_\Ty$, $f_\Sub$, $f_\Tm$ or just write $f$ if it is
clear which one is meant.
\end{definition}

@@

    & (\Gamma^\A:\Set) && \hspace{-0.3em}\times && (\Gamma^\D:\Gamma^\A\ra\Set) && \hspace{-0.3em}\times && (\Gamma^\S:(\gamma:\Gamma^\A)\ra\Gamma^\D\,\gamma\ra\Set) \\

>>>>

Inductive-inductive types (IITs \cite{forsberg-phd}) allow the mutual
definition of a type and a family of types over the first one. IITs
were originally introduced to represent the well-typed syntax of type
theory itself, and the main example is still the Chapman-style
\cite{chapman09eatitself} syntax for a type theory. A minimised version
is the IIT of contexts and types given by the following constructors.
\begin{alignat*}{5}
  & \Con && : \Set \\
  & \Ty && : \Con\ra\Set \\
  & \mathsf{empty} && : \Con \\
  & \mathsf{ext} && : (\GG:\Con)\ra\Ty\,\GG\ra\Con \\
  & \mathsf{U} && : (\GG:\Con)\ra\Ty\,\GG \\
  & \mathsf{El} && : (\GG:\Con)\ra\Ty\,(\mathsf{ext}\,\GG\,(\mathsf{U}\,\GG))
%  & \mathsf{Pi} && : (\GG:\Con)(A:\Ty\,\GG)\ra\Ty\,(\mathsf{ext}\,\GG\,A)\ra\,\Ty\,\GG
\end{alignat*}
This type has two sorts, $\Con$ and $\Ty$. The $\mathsf{ext}$ constructor of
$\Con$ refers to $\Ty$ and the $\Ty$-constructor $\mathsf{U}$ refers to $\Con$,
hence the two sorts have to be defined simultaneously. Moreover, $\Ty$ is
indexed over $\Con$. This precludes a reduction analogous to the reduction of
$\mathsf{isEven}$--$\mathsf{isOdd}$, as we would get a type indexed over
itself. Another unique feature of IITs (which also holds for higher inductive
types \cite{HoTTbook}) is that later constructors can refer to previous
constructors: in our case, $\mathsf{El}$ mentions
$\mathsf{ext}$.
% \[
%   \mathsf{Con?} : (b:\mathsf{Bool})\ra\mathsf{if}\,b\,\mathsf{then}\,\Set\,\mathsf{else}\,(\mathsf{Con?}\,\mathsf{false}\ra\Set),
% \]

The elimination principle for the above IIT has the following two
motives (one for each sort) and four methods (one for each
constructor).
\begin{alignat*}{5}
  & Con^D && : \Con\ra\Set \\
  & Ty^D && : Con^D\,\GG\ra\Ty\,\GG\ra\Set \\
  & {empty}^D && : Con^D\,\mathsf{empty} \\
  & {ext}^D && : (\GG^D:Con^D\,\GG)\ra Ty^D\,\GG^D\,A\ra Con^D\,(\mathsf{ext}\,\GG\,A) \\
  & U^D && : (\GG^D:Con^D\,\GG)\ra Ty^D\,\GG^D\,(\mathsf{U}\,\GG) \\
  & El^D && : (\GG^D:Con^D\,\GG)\ra\,Ty^D\,({ext}^D\,\GG^D\,(U^D\,\GG^D))\,(\mathsf{El}\,\GG)
%  & Pi^D && : (\GG^D:Con^D\,\GG)(A^D:\Ty^D\,\GG^D\,A)\ra \Ty^D\,(\mathsf{ext}^D\,\GG^D\,A^D)\ra\,\Ty^D\,\GG^D\,(\mathsf{Pi}\,\GG\,A\,B)
\end{alignat*} xxx
Above we used implicit quantifications for $\GG:\Con$ and
$A:\Ty\,\GG$ to ease readability, e.g.\ $Ty^D$ has an implicit
parameter $\GG$ before its explicit parameter of type
$Con^D\,\GG$.

Given the above motives and methods the elimination principle provides
two functions
\begin{alignat*}{5}
  & \mathsf{elimCon} && : (\GG:\Con)\ra Con^D\,\GG \\
  & \mathsf{elimTy} && : (A:\Ty\,\GG)\ra Ty^D\,(\mathsf{elimCon}\,\GG)\,A
\end{alignat*}
with the following computation rules.
\begin{alignat*}{5}
  & \mathsf{elimCon}\,\mathsf{empty} && = {empty}^D \\
  & \mathsf{elimCon}\,(\mathsf{ext}\,\GG\,A) && = {ext}^D\,(\mathsf{elimCon}\,\GG)\,(\mathsf{elimTy}\,A) \\
  & \mathsf{elimTy}\,(\mathsf{U}\,\GG) && = U^D\,(\mathsf{elimCon}\,\GG) \\
  & \mathsf{elimTy}\,(\mathsf{El}\,\GG) && = El^D\,(\mathsf{elimCon}\,\GG)
\end{alignat*}
The functions $\mathsf{elimCon}$ and $\mathsf{elimTy}$ are an example
of a \emph{recursive-recursive} definition (using nomenclature from
\cite{forsberg-phd}). This means two mutually defined functions where
the type of the second function depends on the first function. The
proof assistant Agda \cite{norell07thesis} allows defining such
functions (even from non-IITs) and it is currently the only proof
assistant supporting IITs\footnote{An experimental version of Coq with
  IITs is also available on GitHub.}.

Reducing IITs to inductive types (more precisely, to indexed W-types)
is an open problem. Forsberg \cite{forsberg-phd} showed a reduction in
extensional type theory however this only provides a simpler,
non-recursive-recursive elimination principle. Hugunin \cite{jasper}
reduced several IITs to inductive types, working inside a cubical type
theory, but he also only constructed the simple eliminator. To
illustrate the difference, we list the motives and methods and the
simple elimination principle for the $\Con$--$\Ty$ example. Again, we
use implicit quantifications.
\begin{alignat*}{5}
  & Con^S && : \Con\ra\Set \\
  & Ty^S && : \Ty\,\GG\ra\Set \\
  & {empty}^S && : Con^S\,\mathsf{empty} \\
  & {ext}^S && : Con^S\,\GG\ra Ty^S\,\,A\ra Con^S\,(\mathsf{ext}\,\GG\,A) \\
  & U^S && : Con^S\,\GG\ra Ty^S\,(\mathsf{U}\,\GG) \\
  & El^S && : Con^S\,\GG\ra\,Ty^S\,(\mathsf{El}\,\GG) \\
  & \mathsf{selimCon} && : (\GG:\Con)\ra Con^S\,\GG \\
  & \mathsf{selimTy} && : (A:\Ty\,\GG)\ra Ty^S\,A
\end{alignat*}
This simple elimination principle is too weak e.g.\ to define the
standard (metacircular) interpretation of our small syntax
\cite{ttintt}. Using pattern matching notation, this interpretation is
the following:
\begin{alignat*}{5}
  & \ll\blank\rr && : \Con\ra\Set_1 \\
  & \ll\blank\rr && : \ll\GG\rr\ra\Set_1 \\
  & \ll\mathsf{empty}\rr && := \top \\
  & \ll\mathsf{ext}\,\GG\,A\rr && := (\gamma:\ll\GG\rr)\times\ll A\rr\,\gamma \\
  & \ll\mathsf{U}\,\GG\rr\,\gamma && := \Set \\
  & \ll\mathsf{El}\,\GG\rr\,(\gamma,X) && := X
\end{alignat*}
The reason that we need the general elimination principle is that
$\ll\blank\rr$ for types refers to $\ll\blank\rr$ for contexts, hence
this function is recursive-recursive.

@@

Just as ibid., we only consider finitary IITs, that is, constructors
can only have a finite number of recursive arguments. An example
constructor for $\Con$--$\Ty$ which is not allowed is the following:
\[
  \Pi_\infty : (\GG : \Con) \rightarrow (\mathbb{N} \rightarrow \Ty\,\GG) \rightarrow \Ty\,\GG
\]

@@

\begin{definition}[Signature algebra, $\SignAlg$]\label{def:algebra}
  In a model of ETT, a signature algebra is an iterated $\Sigma$ type
  consisting of the following four (families of) sets, 17 operations
  and 18 equalities.
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\
  & \Con && : \Set \\
  & \Ty  && : \Con\ra\Set \\
  & \Sub  && : \Con\ra\Con\ra\Set \\
  & \Tm  && : (\GG:\Con)\ra\Ty\,\GG\ra\Set \\
  & \id && : \Sub\,\GG\,\GG \\
  & \blank\circ\blank && : \Sub\,\GT\,\GD\ra\Sub\,\GG\,\GT\ra\Sub\,\GG\,\GD \\
  & \ass && : (\sigma \circ \delta) \circ \nu = \sigma \circ (\delta \circ \nu) \\
  & \idl && : \id\circ\sigma = \sigma \\
  & \idr && : \sigma\circ\id = \sigma \\
  & \blank[\blank] && : \Ty\,\GD\ra\Sub\,\GG\,\GD\ra\Ty\,\GG \\
  & \blank[\blank] && : \Tm\,\GD\,A\ra(\sigma:\Sub\,\GG\,\GD)\ra\Tm\,\GG\,(A[\sigma]) \\
  & [\id] && : A[\id] = A \\
  & [\circ] && : A[\sigma \circ \delta] = A[\sigma][\delta] \\
  & [\id] && : t[\id] = t \\
  & [\circ] && : t[\sigma \circ \delta] = t[\sigma][\delta] \\
  & \boldsymbol{\cdot} && : \Con \\
  & \epsilon && : \Sub\,\GG\,\boldsymbol{\cdot} \\
  & {\boldsymbol{\cdot}\eta} && : (\sigma : \Sub\,\GG\,\boldsymbol{\cdot}) \ra \sigma = \epsilon \\
  & \blank\ext\blank && : (\GG:\Con)\ra\Ty\,\GG\ra\Con \\
  & \blank,\blank && : (\sigma:\Sub\,\GG\,\GD)\ra\Tm\,\GG\,(A[\sigma])\ra\Sub\,\GG\,(\GD\ext A) \\
  & \pi_1 && : \Sub\,\GG\,(\GD\ext A)\ra \Sub\,\GG\,\GD \\
  & \pi_2 && : (\sigma : \Sub\,\GG\,(\GD\ext A))\ra \Tm\,\GG\,(A[\pi_1 \sigma]) \\
  & {\pi_1\beta} && : \pi_1(\sigma, t) = \sigma \\
  & {\pi_2\beta} && : \pi_2(\sigma, t) = t \\
  & {\pi\eta} && : (\pi_1\,\sigma, \pi_2\,\sigma) = \sigma \\
  & {,\circ} && : (\sigma, t)\circ\delta = (\sigma\circ\delta, t[\delta]) \\
  & \rlap{(2) Universe} \\
  & \U && : \Ty\,\GG \\
  & \El && : \Tm\,\GG\,\U\ra\Ty\,\GG \\
  & {\U[]} && : \U[\sigma] = \U \\
  & {\El[]} && : (\El\,a)[\sigma] = \El\,(a[\sigma]) \\
  & \rlap{(3) Inductive parameters} \\
  & \Pi && : (a:\Tm\,\GG\,\U)\ra\Ty\,(\GG\ext\El\,a)\ra\Ty\,\GG \\
  & \blank\oldapp\blank && : \Tm\,\GG\,(\Pi\,a\,B)\ra (u : \Tm\,\GG\,(\El\,a))
  \ra \Tm\,\GG\,(\El\,(B[\id,\,u])) \\
  & {\Pi[]} && : (\Pi\,a\,B)[\sigma] = \Pi\,(a[\sigma])\,(B[\sigma\circ\p,\q]) \\
  & {\oldapp[]} && : (t\oldapp\alpha)[\sigma] = (t[\sigma])\mathop{\oldapp}(\alpha[\sigma]) \\
  & \rlap{(4) External parameters} \\
  & \Pim && : (T:\Set)\ra(T\ra\Ty\,\GG)\ra\Ty\,\GG \\
  & \blank\appm \blank && : \Tm\,\GG\,(\Pim\,T\,B)\ra(\alpha:T) \ra\Tm\,\GG\,(B\,\alpha) \\
  & {\Pim[]} && : (\Pim\,T\,B)[\sigma] = \Pim\,T\,(\lambda \alpha.(B\,\alpha)[\sigma]) \\
  & {\appm[]} && : (t\appm\alpha)[\sigma] = (t[\sigma])\mathop{\appm}\alpha
\end{alignat*}
Given an $M:\SignAlg$, we denote its components by $\Con^M$, $\Ty^M$,
$\Sub^M$, $\Tm^M$, $\id^M$, and so on. We omit the indices if there is
only one signature algebra in scope (e.g.\ in
\Cref{def:abbrevs,ex:contexts}).
\end{definition}

\begin{definition}[Abbreviations]\label{def:abbrevs}
  For a signature algebra, we use $\wk : \Sub\,(\GG\ext A)\,\GG$
  to mean $\pi_1\,\id$. We recover de Bruijn indices by setting
  $0:= \pi_2 \,\id$ and $1+n := n[\wk]$. $\Pi\,a\,(B[\wk])$ is
  abbreviated by $a\Ra B$, $\Pim\,T\,(\lambda\_.B)$ by $T\Ram B$.
\end{definition}

@@

  The $\Con$--$\Ty$ example from \Cref{sec:intro} is given by the
  following context.
  \begin{alignat*}{5}
    & \boldsymbol{\cdot} \,\ext                                                 && \boldsymbol{\cdot} \,\ext                                                                \\
    & \U\,\ext                                                                  && Con && : \U\,\ext                                                                        \\
    & 0\Ra\U\,\ext                                                              && Ty && : Con\Ra\U\,\ext                                                                   \\
    & \El\,1\,\ext                                                              && empty && : \El\,Con\,\ext                                                                \\
    & \Pi\,2\,(2\oldapp 0 \Ra \El\,3)\ext                                       && ext && : \Pi\,(\GG : Con)\,(Ty\oldapp\GG \Ra \El\,Con)\ext                         \\
    & \Pi\,3\,(\El\,(3\oldapp 0))\,\ext                                         && U && : \Pi\,(\GG:Con)\,(\El\,(Ty\oldapp\GG))\,\ext                                 \\
    & \Pi\,4\,(\El\,(4\oldapp(2\oldapp 0\oldapp (1\oldapp 0)))) \hspace{4.79em} && El && : \Pi\,(\GG : Con)\,(\El\,(Ty\oldapp(ext\oldapp\GG\oldapp (U\oldapp\GG))))
  \end{alignat*}
\end{example}

@@

The notion of morphism is determined by the notion of algebra
\cite{large_inf_qiit}, but we include it here for completeness.
\begin{definition}[Signature morphism, $\SignMor$]\label{def:morphism}
  A morphism from signature algebras $M$ to $N$ denoted
  $\SignMor\,M\,N$ consists of four functions and 17 equalities
  expressing that the functions preserve the operations of the two
  algebras. We use the same naming as in \Cref{def:algebra} and use
  superscripts to denote which algebra is meant.
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\
  & \Con && : \Con^M && \ra\Con^N \\
  & \Ty  && : \Ty^M\,\GG && \ra\Ty^N\,(\Con\,\GG) \\
  & \Sub  && : \Sub^M\,\GG\,\GD && \ra\Sub^N(\Con\,\GG)\,(\Con\,\GD) \\
  & \Tm  && : \Tm^M\,\GG\,A && \ra\Tm^N\,(\Con\,\GG)\,(\Ty\,A) \\
  & \id && : \Sub\,\id^M && = \id^N \\
  & {\circ} && : \sigma\circ^M\delta && = \Sub\,\sigma\circ^N\Sub\,\delta \\
  & [] && : A[\sigma]^M && = \Ty\,A[\Sub\,\sigma]^N \\
  & [] && : t[\sigma]^M && = \Tm\,t[\Sub\,\sigma]^N \\
  & \boldsymbol{\cdot} && : \Con\,\boldsymbol{\cdot}^M && = \boldsymbol{\cdot}^N \\
  & \epsilon && : \Sub\,\epsilon^M && = \epsilon^N \\
  & \ext && : \Con\,(\GG\ext^M A) && = \Con\,\GG\ext^N\Ty\,A \\
  & , && : \Sub\,(\sigma,^M t) && = \Sub\,\sigma ,^N\Tm\,t \\
  & \pi_1 && : \Sub\,(\pi_1^M\,\sigma) && = \pi_1^N\,(\Sub\,\sigma) \\
  & \pi_2 && : \Tm\,(\pi_2^M\,\sigma) && = \pi_2^N\,(\Sub\,\sigma) \\
  & \rlap{(2) Universe} \\
  & \U && : \Ty\,\U^M && = \U^N \\
  & \El && : \Ty\,(\El^M\,a) && = \El^N\,(\Tm\,a) \\
  & \rlap{(3) Inductive parameters} \\
  & \Pi && : \Ty\,(\Pi^M\,a\,B) && = \Pi^N\,(\Tm\,a)\,(\Ty\,B) \\
  & \oldapp && : \Tm\,(t\mathbin{{\oldapp}^M} u) && = \Tm\,t\mathbin{{\oldapp}^N}\,\Tm\,u \\
  & \rlap{(4) External parameters} \\
  & \Pim && : \Ty\,(\Pim^M\,T\,B) && = \Pim^N\,T\,(\lambda\alpha.\Ty\,(B\,\alpha)) \\
  & \appm && : \Tm\,(t\mathbin{{\appm}^M}\alpha) && = \Tm\,t\mathbin{{\appm}^N}\alpha
\end{alignat*}
Given an $f:\SignMor\,M\,N$, we denote its first four components just
by $f_\Con$, $f_\Ty$, $f_\Sub$, $f_\Tm$ or just write $f$ if it is
clear which one is meant.
\end{definition}

@@

    & (\GG^\A:\Set) && \hspace{-0.3em}\times && (\GG^\D:\GG^\A\ra\Set) && \hspace{-0.3em}\times && (\GG^\S:(\gamma:\GG^\A)\ra\GG^\D\,\gamma\ra\Set) \\
    & (A^\A:\GG^\A\ra\Set) && \hspace{-0.3em}\times && (A^\D:\GG^\D\,\gamma\ra A^\A\,\gamma\ra\Set) && \hspace{-0.3em}\times && (A^\S:\GG^\S\,\gamma\,\gamma^D\ra(\alpha:A^\A\,\gamma)\ra \\
